Static contract checking for Haskell

### 2011-11-06 paper.tex
Paper: removed $\Phi$, the union of formulas, and replaced with conjunction.

We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
  CF(fst(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
can't think of a good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons(y,ys) \neq \unr 
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\Phi_4$) say that constructors are ``lazy'' or ``total''.}

Finally, we define projections for $Cons$. \annote[nc]{It is not
  strictly necessary, but it will be handy:}{??? Why is it handy? It
  doesn't seem to play any role in formal part of this paper (besides
  being specified in axioms $\Phi_1$ (projections)). In the current
  implementation it's used to avoid introducing existentially
  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
  functions defined by case expressions. Equation \eqref{unrnotnull}
  below uses it that way. ??? Does Equinox care?}
\begin{equation*}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}\note[nc]{??? Why not the simpler \[\forall y,ys.~sel_{2,Cons}(Cons(y,ys)) = ys\]? Same question in Figure \ref{dtrans}.  (Equinox does have a built in notion of equality, e.g. running equinox on
\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
 results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
%% \[
%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
%%                  \text{\texttt{| equinox /dev/stdin}}}}
%% \]

@@

\begin{figure}
 \centering
  \[  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr \\
    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array} \]
  \caption{First-order logic syntax}
  \note[nc]{The $\Phi$ are unnecessary and don't make things simpler, because a set of formulas is equivalent to the conjunction of its contents.  XXX, TODO: drop $\Phi$ and replace all $\bigcup$ with $\bigwedge$.}
  \label{fol-stx}
\end{figure}

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

@@

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = K_1(\vec{x_1}) &\to& a = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = K_n(\vec{x_n}) &\to& a = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

\subsection{$\ktrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ No full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
\end{itemize}

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ctrans{contract} = \Phi} $$
\begin{eqnarray*}
  \ctrans{e \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = true \lor e = \unr)\\
  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{e~x} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
  \ctrans{e \in \cfc} &=& \min{e} \land \cf{e}
\end{eqnarray*}

>>>>

We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
  CF(fst(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
can't think of a good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons(y,ys) \neq \unr 
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phi_\text{lazy}$) say that constructors are ``lazy'' or ``total''.}

Finally, we define projections for $Cons$. \annote[nc]{It is not
  strictly necessary, but it will be handy:}{??? Why is it handy? It
  doesn't seem to play any role in formal part of this paper (besides
  being specified in axioms $\phi_\text{projection}$). In the current
  implementation it's used to avoid introducing existentially
  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
  functions defined by case expressions. Equation \eqref{unrnotnull}
  below uses it that way. ??? Does Equinox care?}
\begin{equation*}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}\note[nc]{??? Why not the simpler \[\forall y,ys.~sel_{2,Cons}(Cons(y,ys)) = ys\]? Same question in Figure \ref{dtrans}.  (Equinox does have a built in notion of equality, e.g. running equinox on
\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
 results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
%% \[
%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
%%                  \text{\texttt{| equinox /dev/stdin}}}}
%% \]

@@

\begin{figure}
 \centering
  \[  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr \\
    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
  \end{array} \]
  \caption{First-order logic syntax}
  \label{fol-stx}
\end{figure}

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \phi~(\mbox{formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formulae}) \\
    \ctrans{f \in c} &\to& \phi~(\mbox{formulae}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

@@

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = K_1(\vec{x_1}) &\to& a = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = K_n(\vec{x_n}) &\to& a = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

\subsection{$\ktrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$

\begin{itemize}
\item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\phi_\text{disjoint})$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\phi_\text{lazy})$ No full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
\end{itemize}

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{y})) \\
      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

\begin{figure*}
  $$ \ruleform{\ctrans{contract} = \phi} $$
\begin{eqnarray*}
  \ctrans{e \in \{x \mid b \}} &=&  \min{b[e/x]} \land (b[e/x] = true \lor e = \unr)\\
  \ctrans{e \in x:c_1 \to c_2} &=& \forall x.~\min{e~x} \to (\ctrans{x \not \in c_1} \lor  \ctrans{e~x \in c_2})\\
  \ctrans{e \in \cfc} &=& \min{e} \land \cf{e}
\end{eqnarray*}
